Tail Flexibility in the Degrees of Preferential Attachment Networks

### 2025-05-07 paper.tex
make equations & figures larger in qmd

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Tail Flexibility in the Degrees of Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, extremes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

% keywords
\keywords{networks,  extremes}

@@

\subsection{Structure of this paper}\label{structure-of-this-paper}

Section~\ref{sec-gpa} looks at the theoretical limiting behaviour of a
GPA model when \(m=1\) according to the results from \citep{rudas07},
going on to introduce a flexible class of preference functions that can
guarantee a heavy tailed degree distribution while still being flexible
in the body. Section~\ref{sec-rec} shows that the preference function
parameters can be fairly well recovered from the degree distributions of
networks simulated from the GPA model using various functions of the
class introduced in Section~\ref{sec-gpa}. Having shown that for
simulated data the model parameters are recoverable from the degree
distribution alone, Section~\ref{sec-real} builds upon this and attempts
to fit the model to degree distributions of real networks; modelling not
only the degree distribution itself but also the estimates for the
parameters of the preference function assuming the networks evolved
according to the GPA scheme. Ultimately, Section~\ref{sec-conc}
discusses the main results of the paper, pitfalls of this method and
future work.

@@

which aligns with the result from \citep{rudas07}, where the expression
is derived using the same branching process results.

Using results from \citep{shimura12} the domain of attraction of this
limiting degree distribution can be found.

@@

Previously, in \citep{krapivsky01}, a preference function of the form
\(k^\alpha +\varepsilon\) has been considered. However, it has been
noted that super-linear preferential attachment (\(\alpha > 1\)) leads
to a degenerate limiting degree distribution where at some point one
vertex gains the connection from every vertex that joins the network,
additionally sub-linear preferential attachment (\(\alpha <1\)) has been
shown to lead to a light tailed limiting degree distribution. Consider a
preference function of the form:

@@

\begin{equation}\phantomsection\label{eq-polysurv}{
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k\ge k_0.
\end{cases}
}\end{equation}

with \(\lambda^*\) satisfying \(\hat \rho(\lambda^*)=1\) where:

\[
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon} 
\]

which must be solved numerically for most parameter choices. Also, note
that \(\lambda>\beta\) since when \(\lambda\le \beta\) the infinite sum
no longer converges and instead goes to infinity.

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-polylinsurv-1.pdf}

@@

The survival (Equation~\ref{eq-polysurv}) can be connected to the IGPD
mentioned in Section~\ref{sec-intro} by using Sterling's approximation
of the gamma function to obtain:

\[
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
\] meaning that for \(k\ge k_0\) the limiting degree distribution is
similar to
\(\text{IGPD}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)\).

To assess how close of an approximation this is the theoretical
conditional survivals are shown in Figure~\ref{fig-approx_surv} in
colour and their IGPD approximations are shown in grey. The
approximation seems to hold up fairly well even for large degrees.

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-approx_surv-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-polyheat-1.pdf}

@@

\caption{\label{fig-polyheat}Heat maps of \(\xi\) for various
combinations of the parameters of the proposed model.}

@@

\begin{equation}\phantomsection\label{eq-linsurv}{
\bar F(k) = \begin{cases}
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k+1+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k+1+\lambda^*+\epsilon)},&k<k_0\\
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times \frac{\Gamma\left(\frac{k_0+\varepsilon+\lambda^*}{\beta}\right)}{\Gamma\left(\frac{k_0+\varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon}{\beta}+1\right)}{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon+\lambda^*}{\beta}+1\right)},&k\ge k_0
\end{cases}
}\end{equation}

Just like the previous model, the survival can be approximated by:

\[
\bar F(k) \approx \begin{cases}
\frac{\Gamma(\lambda^*+\varepsilon)}{\Gamma(\varepsilon)} (1+\varepsilon)^{\lambda^*}\left(\frac{k}{1+\varepsilon} + 1\right)^{-\lambda^*},&k<k_0\\
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times\left(\frac{\beta(k+1-k_0)}{k_0+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0
\end{cases}
\]

@@

\section{Recovery}\label{sec-rec}

The goal of was to provide a possible preference function that is able
to explain the growth of real networks with varied tail behaviour in
their degree distributions. This section aims to show that the
parameters of the model in Section~\ref{sec-model} can be recovered from
the degree distribution of a network simulated from it.

>>>>

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Tail Flexibility in the Degrees of Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, discrete extremes, degree distribution},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

% keywords
\keywords{networks,  discrete extremes,  degree distribution}

@@

The rest of this paper is as follows: Section~\ref{sec-gpa} looks at the
theoretical limiting behaviour of a GPA model when \(m=1\) according to
the results from \citep{rudas07}, going on to introduce a flexible class
of preference functions that can guarantee a heavy tailed degree
distribution while still being flexible in the body.
Section~\ref{sec-rec} shows that the preference function parameters can
be fairly well recovered from the degree distributions of networks
simulated from the GPA model using various functions of the class
introduced in Section~\ref{sec-gpa}. Having shown that for simulated
data the model parameters are recoverable from the degree distribution
alone, Section~\ref{sec-real} builds upon this and attempts to fit the
model to degree distributions of real networks; modelling not only the
degree distribution itself but also the estimates for the parameters of
the preference function assuming the networks evolved according to the
GPA scheme. Ultimately, Section~\ref{sec-conc} discusses the main
results of the paper, pitfalls of this method and future work.

@@

which aligns with the result from \citet{rudas07}, where the expression
is derived using the same branching process results.

Using results from \citet{shimura12}, the domain of attraction of this
limiting degree distribution can be found.

@@

Previously, in \citet{krapivsky01}, a preference function of the form
\(k^\alpha +\varepsilon\) has been considered. However, it has been
noted that super-linear preferential attachment (\(\alpha > 1\)) leads
to a degenerate limiting degree distribution where at some point one
vertex gains the connection from every vertex that joins the network,
additionally sub-linear preferential attachment (\(\alpha <1\)) has been
shown to lead to a light tailed limiting degree distribution. Consider a
preference function of the form:

@@

\begin{equation}\phantomsection\label{eq-polysurv}{
\bar F(k) = \begin{cases}
\displaystyle\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0\\
\displaystyle\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k\ge k_0,
\end{cases}
}\end{equation}

with \(\lambda^*\) satisfying

\[
\hat\rho(\lambda^{*}) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda^{*}+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda^{*}-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^{*} + i^\alpha + \varepsilon} = 1,
\]

which has to be solved numerically for most parameter choices. Note that
\(\lambda>\beta\) since when \(\lambda\le \beta\) the infinite sum no
longer converges and instead goes to infinity.

@@

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-polylinsurv-1.pdf}

@@

Applying Stirling's approximation of the gamma function to the second
case of Equation~\ref{eq-polysurv}, we obtain, for \(k\geq k_0\),

\[
\bar F(k) 
%\begin{cases}
%=\displaystyle\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\displaystyle\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},%&k\ge k_0,
%\end{cases}
\] which is the survival function of the
\(\text{IGPD}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)\)
distribution. This means that, above the threshold of the piecewise
preference function, the limiting degree distribution can be
approximated by the IGPD with shape parameter \(\beta/\lambda^{*}\).

To assess the quality of the approximation, the conditional survivals
are shown in Figure~\ref{fig-approx_surv} in colour and their IGPD
approximations are shown in grey. The approximation seems to hold up
fairly well even for large degrees.

@@

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-approx_surv-1.pdf}

@@

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-polyheat-1.pdf}

@@

\caption{\label{fig-polyheat}Heatmaps of \(\xi\) for various
combinations of the parameters of the proposed model.}

@@

\begin{equation}\phantomsection\label{eq-linsurv}{
\bar F(k) = \begin{cases}
\displaystyle\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k+1+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k+1+\lambda^*+\epsilon)},&k<k_0,\\
\displaystyle\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times \frac{\Gamma\left(\frac{k_0+\varepsilon+\lambda^*}{\beta}\right)}{\Gamma\left(\frac{k_0+\varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon}{\beta}+1\right)}{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon+\lambda^*}{\beta}+1\right)},&k\ge k_0.
\end{cases}
}\end{equation}

Similar to the general case, the survival function can be approximated
by:

\[
\bar F(k) \approx \begin{cases}
\displaystyle\frac{\Gamma(\lambda^*+\varepsilon)}{\Gamma(\varepsilon)} (1+\varepsilon)^{\lambda^*}\left(\frac{k}{1+\varepsilon} + 1\right)^{-\lambda^*},&k<k_0,\\
\displaystyle\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times\left(\frac{\beta(k+1-k_0)}{k_0+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0.
\end{cases}
\]

@@

\section{Simulation study}\label{sec-rec}

This section aims to show that the parameters of the model in
Section~\ref{sec-model} can be recovered from the degree distribution of
a network simulated from it.
